Generalized Universe Hierarchies and First-Class Universe Levels

### 2021-02-21 paper.tex
progress

\begin{abstract}
In type theories, universe hierarchies are commonly used to increase the
expressive power of the theory while avoiding inconsistencies arising from size
issues. There are numerous ways to specify universe hierarchies, and theories
may differ in details of cumulativity, choice of universe levels, specification
of type formers and eliminators, and available internal operations on levels. In
the current work, we aim to provide a framework which covers a large part of the
design space. First, we develop syntax and semantics for cumulative universe
hierarchies, where levels may come from any set equipped with a transitive
well-founded ordering. In the semantics, we show that induction-recursion can be
used to model transfinite hierarchies, and also support lifting operations on
type codes which strictly preserve type formers. Then, we consider a setup where
universe levels are first-class types and subject to arbitrary internal
reasoning. This generalizes the bounded universe polymorphism features of Coq
and also the internal level computations in Agda.
\end{abstract}

@@

\todo{purge unused features}

We work in a Martin-Löf type theory which has the following features.
\begin{itemize}
  \item Two universes named $\Set0$ and $\Set1$, where $\Set0$ supports
    inductive-recursive types (IR) as specified by Dybjer and Setzer
    \cite{dybjer99finite}. The $\Set1$ universe is not essential and we only use
    it as a convenience feature, in this paper and in the Agda formalization. We
    may omit the universe indices if they can be inferred or if we work over
    arbitrary indices.
  \item Function extensionality and uniqueness of identity proofs
    (UIP). Additionally, we assume equality reflection in this paper, thus
    working in extensional type theory, to avoid noise from equality transports.
  \item We write function types as $(x : A)\to B$ with $\lambda\,x.\,t$
    inhabitants, and $\Sigma$-types as $(x : A) \times B$, with pairing as
    $(t,\, u)$ and projections as $t.1$ and $t.2$. We have $\top$ as the unit
    type with inhabitant $\tt$, and $\bot$ as the empty type. Propositional identity
    is written as $t = u$ with $\refl : t = t$.
  \item We occasionally use $\{x : A\} \to B$ for an Agda-like notation for
    function types with implicit arguments, and omit implicit applications
    whenever they can be inferred. We may omit implicit function types
    altogether if it is clear where certain variables are quantified.
\end{itemize}

@@

\begin{mydefinition}
A \emph{category with family} (cwf) \cite{Dybjer96internaltype} consists of the following data:
\begin{itemize}
\item A category with a terminal object. We denote the set of objects as $\Con :
  \Seti$ and use capital Greek letters starting from $\Gamma$ to refer to
  objects. The set of morphisms is $\Sub : \Con \to \Con \to \Seti$, and we use
  $\sigma$, $\delta$ and so on to refer to morphisms. The terminal object is
  $\emptycon$ with unique morphism $\epsilon : \Sub\,\Gamma\,\emptycon$. In
  initial models (that is, syntaxes) of type theories, objects correspond to
  typing contexts, morphisms to parallel substitutions and the terminal object to
  the empty context; this informs the naming scheme that we use here.
\item A \emph{family structure}, containing $\Ty : \Con \to \Seti$ and $\Tm :
  (\Gamma : \Con) \to \Ty\,\Gamma \to \Seti$, where $\Ty$ is a presheaf over the
  category of contexts and $\Tm$ is a presheaf over the category of elements of
  $\Ty$. This means that both types ($\Ty$) and terms ($\Tm$) can be
  substituted, and substitution has functorial action. We use $A$, $B$, $C$ to
  refer to types and $t$, $u$, $v$ to refer to terms, and use $A[\sigma]$ and
  $t[\sigma]$ for substituting types and terms. Additionally, a family structure
  has \emph{context comprehension} which consists of an context extension
  operation $\blank\ext\blank : (\Gamma : \Con) \to \Ty\,\Gamma \to \Con$
  together with an isomorphism $\Sub\,\Gamma\,(\Delta\ext A) \simeq ((\sigma :
  \Sub\,\Gamma\,\Delta) \times \Tm\,\Gamma\,(A[\sigma]))$ which is natural in
  $\Gamma$.
\end{itemize}
\end{mydefinition}

@@

There are other ways for presenting the basic categorical structure of models,
which are nonetheless equivalent to cwfs, including natural models \cite{awodey18natural}
and categories with attributes \cite{cartmellthesis}. We use the cwf presentation for its
immediately algebraic character and closeness to syntax, as an explicit
substitution calculus can be directly read off the specification.

@@

In the following, whenever we introduce a type former, we always assume that it
is natural with respect to substitution, i.e.\ all type and term formers have a
corresponding substitution rule.

@@

\todo{Mention dependent right adjoints}

@@

Assume an inclusion from ($\Ty_0$, $\Tm_0$) to ($\Ty_1$, $\Tm_0$). Now, we can
eliminate from $\Bool_0$ to $\Bool_1$. If we have some $b :
\Tm_0\,\Gamma\,\Bool_0$, we also have $\up\!b :
\Tm_1\,\Gamma\,(\Lift_1\,\Bool_0)$, hence $\up\!b :
\Tm_1\,\Gamma\,\Bool_1$. Then, we can use $\Bool_1$ elimination, as in
$\mathsf{if}\,\up\!b\,\,\mathsf{then}\,\true_1\,\mathsf{else}\,\false_1 :
\Tm_1\,\Gamma\,\Bool_1$. The $\up$ computation ensures that the eliminator
computes appropriately on canonical terms: if $b$ is $\true_0$, then we get
$\up\!\true_0 = \true_1$ as the if-then-else scrutinee.

@@

A strict inclusion corresponds to Sterling's \emph{algebraic cumulativity}
\cite{sterling2019algebraic}. The additional equations are a matter of convenience: they allow us
to omit term liftings in informal syntax\footnote{In a proof assistant, we would
  still have to explicitly transport along the strict inclusion
  equations.}. Most of the time we can also omit level annotations
on term formers. For example, we have $\true_0 : \Tm_0\,\Gamma\,\Bool_0$,
but also $\true_0 : \Tm_0\,\Gamma\,(\Lift\,\Bool_0)$, hence $\true_0 :
\Tm_0\,\Gamma\,\Bool_1$. Moreover, $\true_0$ is definitionally equal to
$\true_1$, since $\true_0 =\,\up\!\true_0 = \true_1$. Thus, using simply $\true$
is fine whenever the family is clear from context.

@@

\begin{mylemma}
Every well-founded relation is irreflexive \cite[Lemma 10.4.2]{hottbook}.
\end{mylemma}

@@

\todo{find reference for Coquand univ}

@@

\begin{mydefinition}[Level reflection rules]
Assume that we have \emph{definitions for internal levels} in the syntax of TTDL (i.e.\
the initial TTDL model):
\begin{alignat*}{3}
  &\Lvl^I               &&: \Ty\,\Gamma\,l_0\\
  &l_0^I,\,l_1^I         &&: \Tm\,\Gamma\,\Lvl^I \\
  &\blank\!<^I\!\blank  &&: \Tm\,\Gamma\,\Lvl^I \to \Tm\,\Gamma\,\Lvl^I \to \Ty\,\Gamma\,l_0\\
  &l_{01}^{I}            &&: \Tm\,\Gamma\,(l_0^I <^I l_1^I)
\end{alignat*}
A \emph{reflection rule} for the above consists of
\begin{enumerate}
  \item $\mkLvl : \Tm\,\Gamma\,\Lvl^I \to \Lvl\,\Gamma$ with its inverse $\unLvl$, such that $\mkLvl\,l_0^I = l_0$ and $\mkLvl\,l_1^I = l_1$.
  \item $\mkMor\,\,: \Tm\,\Gamma\,(i <^I j) \to \mkLvl\,i < \mkLvl\,i$ with its inverse
        $\unMor$.
\end{enumerate}
\end{mydefinition}

For any definition of internal levels, we may extend the specification of TTDL
with the corresponding reflection rule, thereby getting an algebraic signature
for a TTFL. We can easily get a TTFL with an IR model in the following
way. First, pick an external $\Lvl$ structure which a) satisfies the
bootstrapping assumption b) has sets of levels and morphisms which can be
represented with syntactic TTDL types.

@@

In this section we give a tour of some of the features expressible in TTFL.

@@

\textbf{Induction on levels and level morphisms.} In Agda 2.6.1, there is an
internal type of finite levels, and while construction rules and some built-in
operations on levels are exposed, there is no general elimination rule on levels.
Thus, there is a $\Nat \to \Lvl$ function but no non-constant $\Lvl \to \Nat$ one.
In contrast, TTFL supports arbitrary elimination on levels, and in particular levels
can be definitionally isomorphic to natural numbers.

\textbf{Type formers returning in least upper bounds of levels}. It is common in
type theories to allow type formers to have parameter types in different
universe levels, say $i$ and $j$, and return in level $i \sqcup j$. In TTFL,
whenever levels are \emph{trichotomous}, meaning that the ordering and equality of
levels is internally decidable, then $\sqcup$ is a definable operation, and the
``heterogeneous'' type formers are derivable\footnote{A level structure which is trichotonomous and supports \emph{extensionality}, i.e.\ $(\forall i.\, (i < j) \iff (i < k)) \to j = k$, is called a \emph{type-theoretic ordinal}. Assuming excluded middle type-theoretic ordinals are equivalent to classical ordinals \cite[Section 10.3]{hottbook}.}.

\textbf{Coercive cumulative subtyping.} TTFL does not directly support
subtyping. However, given enough type formers, it is possible to simply define
subtyping as an interal inductive relation, and interpret a witness of $A \leq B$ as
a coercion function from $A$ to $B$. Also, it should be possible to model a
theory with cumulative subtyping by translation to coercions in TTFL.

  %% data _≤_ (i : Lvl) (j : Lvl) : U i → U j → U (
  %% U≤ : (p : i ≤ j) → U i ≤ U j
  %% Π≤ : A ≤

%% TTFL does not directly support a
%% cumulative subtyping relation. However, it is possible to model subtyping by
%% translation to coercions in TTFL. For example:
%% \begin{itemize}
%%   \item Given $p : i < j$, we may coerce terms from type $\U\,i\,k\,q$ to $\U\,j\,k\,(p \circ q)$,
%%         using $\Lift\,p$.
%%   \item If there is a coercion from $A$ to $A'$, and

%% \end{itemize}

>>>>

\begin{abstract}
In type theories, universe hierarchies are commonly used to increase the
expressive power of the theory while avoiding inconsistencies arising from size
issues. There are numerous ways to specify universe hierarchies, and theories
may differ in details of cumulativity, choice of universe levels, specification
of type formers and eliminators, and available internal operations on levels. In
the current work, we aim to provide a framework which covers a large part of the
design space. First, we develop syntax and semantics for cumulative universe
hierarchies, where levels may come from any set equipped with a transitive
well-founded ordering. In the semantics, we show that induction-recursion can be
used to model transfinite hierarchies, and also support lifting operations on
type codes which strictly preserve type formers. Then, we consider a setup where
universe levels are first-class types and subject to arbitrary internal
reasoning. This generalizes the bounded polymorphism features of Coq and at the
same time the internal level computations in Agda.
\end{abstract}

@@

We work in a Martin-Löf type theory which has the following features.
\begin{itemize}
  \item Two universes named $\Set0$ and $\Set1$, where $\Set0$ supports
    inductive-recursive types (IR) as specified by Dybjer and Setzer
    \cite{dybjer99finite}. The $\Set1$ universe is not essential and we only use
    it as a convenience feature, in this paper and in the Agda formalization. We
    may omit the universe indices if they can be inferred or if we work over
    arbitrary indices.
  \item Function extensionality and uniqueness of identity proofs
    (UIP). Additionally, we assume equality reflection in this paper, thus
    working in extensional type theory, to avoid noise from equality transports.
  \item We write function types as $(x : A)\to B$ with $\lambda\,x.\,t$
    inhabitants, and $\Sigma$-types as $(x : A) \times B$, with pairing as
    $(t,\, u)$. We have $\top$ as the unit type with inhabitant $\tt$, and
    $\bot$ as the empty type. Propositional identity is written as $t = u$
    (coinciding with definitional equality).
  \item We occasionally use $\{x : A\} \to B$ for an Agda-like notation for
    function types with implicit arguments. We usually omit implicit
    applications but may explicitly write them as $t\,\{u\}$.  We may omit
    implicit function types altogether if it is clear where certain variables
    are quantified.
\end{itemize}

@@

\begin{mydefinition}
A \emph{category with family} (cwf) \cite{Dybjer96internaltype} consists of the following data:
\begin{itemize}
\item A category with a terminal object. We denote the set of objects as $\Con :
  \Seti$ and use capital Greek letters starting from $\Gamma$ to refer to
  objects. The set of morphisms is $\Sub : \Con \to \Con \to \Seti$, and we use
  $\sigma$, $\delta$ and so on to refer to morphisms. The terminal object is
  $\emptycon$ with unique morphism $\epsilon : \Sub\,\Gamma\,\emptycon$. In
  initial models (that is, syntaxes) of type theories, objects correspond to
  typing contexts, morphisms to parallel substitutions and the terminal object to
  the empty context; this informs the naming scheme that we use here.
\item A \emph{family structure}, containing $\Ty : \Con \to \Seti$ and $\Tm :
  (\Gamma : \Con) \to \Ty\,\Gamma \to \Seti$, where $\Ty$ is a presheaf over the
  category of contexts and $\Tm$ is a presheaf over the category of elements of
  $\Ty$. This means that both types ($\Ty$) and terms ($\Tm$) can be
  substituted, and substitution has functorial action. We use $A$, $B$, $C$ to
  refer to types and $t$, $u$, $v$ to refer to terms, and use $A[\sigma]$ and
  $t[\sigma]$ for substituting types and terms. Additionally, a family structure
  has \emph{context comprehension} which consists of a context extension
  operation $\blank\ext\blank : (\Gamma : \Con) \to \Ty\,\Gamma \to \Con$
  together with an isomorphism $\Sub\,\Gamma\,(\Delta\ext A) \simeq ((\sigma :
  \Sub\,\Gamma\,\Delta) \times \Tm\,\Gamma\,(A[\sigma]))$ which is natural in
  $\Gamma$.
\end{itemize}
\end{mydefinition}

@@

There are other ways for presenting the basic categorical structure of models,
which are nonetheless equivalent to cwfs, including natural models
\cite{awodey18natural} and categories with attributes \cite{cartmellthesis}. We
use the cwf presentation for its immediately algebraic character and closeness
to conventional explicit substitution syntax.

@@

In the following, whenever we introduce a type or term former, we always assume
that it is natural with respect to substitution, i.e.\ all type and term formers
have a corresponding substitution rule.

@@

We may think about the relation between \emph{modalities} and liftings.  The
main difference is that we have no structural restrictions on variables and
contexts. More concretely, every $\Lift$ is dependent right adjoint
\cite{dependentrightadjoints} to the identity functor on the base category, as
we have $\Tm\,(\msf{Id}\,\Gamma)\,A \simeq \Tm\,\Gamma\,(\Lift\,A)$. Hence,
every lifting can be viewed as a degenerate modality.

@@

Assume an inclusion from ($\Ty_0$, $\Tm_0$) to ($\Ty_1$, $\Tm_0$). Now, we can
eliminate from $\Bool_0$ to $\Bool_1$. If we have some $b :
\Tm_0\,\Gamma\,\Bool_0$, we also have $\up\!b :
\Tm_1\,\Gamma\,(\Lift_1\,\Bool_0)$, hence $\up\!b :
\Tm_1\,\Gamma\,\Bool_1$. Then, we can use $\Bool_1$ elimination, as in
$\mathsf{if}\,\up\!b\,\,\mathsf{then}\,\true_1\,\mathsf{else}\,\false_1 :
\Tm_1\,\Gamma\,\Bool_1$. The $\up$ computation ensures that the eliminator
computes appropriately on canonical terms: if $b$ is $\true_0$, we get
$\up\!\true_0 = \true_1$ as the if-then-else scrutinee.

@@

A strict inclusion corresponds to Sterling's \emph{algebraic cumulativity}
\cite{sterling2019algebraic}. The additional equations are a matter of convenience: they allow us
to omit term liftings in informal syntax\footnote{In a proof assistant, often we would
  still have to explicitly transport along the strict inclusion
  equations.}. Most of the time we can also omit level annotations
on term formers. For example, we have $\true_0 : \Tm_0\,\Gamma\,\Bool_0$,
but also $\true_0 : \Tm_0\,\Gamma\,(\Lift\,\Bool_0)$, hence $\true_0 :
\Tm_0\,\Gamma\,\Bool_1$. Moreover, $\true_0$ is definitionally equal to
$\true_1$, since $\true_0 =\,\up\!\true_0 = \true_1$. Thus, using simply $\true$
is fine whenever the family is clear from context.

@@

%% \begin{mylemma}
%% Every well-founded relation is irreflexive \cite[Lemma 10.4.2]{hottbook}.
%% \end{mylemma}

@@

@@

\begin{mydefinition}[Level reflection rules]
Assume that we have definitions for internal levels in the syntax of TTDL,
i.e.\ all of the following are defined:
\begin{alignat*}{3}
  &\Lvl^I               &&: \Ty\,\Gamma\,l_0\\
  &l_0^I,\,l_1^I         &&: \Tm\,\Gamma\,\Lvl^I \\
  &\blank\!<^I\!\blank  &&: \Tm\,\Gamma\,\Lvl^I \to \Tm\,\Gamma\,\Lvl^I \to \Ty\,\Gamma\,l_0\\
  &l_{01}^{I}            &&: \Tm\,\Gamma\,(l_0^I <^I l_1^I)
\end{alignat*}
A \emph{reflection rule} for the above consists of
\begin{enumerate}
  \item $\mkLvl : \Tm\,\Gamma\,\Lvl^I \to \Lvl\,\Gamma$ with its inverse $\unLvl$, such that $\mkLvl\,l_0^I = l_0$ and $\mkLvl\,l_1^I = l_1$.
  \item $\mkMor\,\,: \Tm\,\Gamma\,(i <^I j) \to \mkLvl\,i < \mkLvl\,i$ with its inverse
        $\unMor$.
\end{enumerate}
\end{mydefinition}

For any definition of internal levels, we may extend the specification of TTDL
with the corresponding reflection rule, thereby getting an algebraic signature
for a type theory with first-class levels (TTFL). We can easily get a TTFL with
an IR model in the following way. First, pick an external $\Lvl$ structure which
a) satisfies the bootstrapping assumption b) has sets of levels and morphisms
which can be represented with syntactic TTDL types.

@@

We describe some of the features expressible in TTFL.

@@

\textbf{Induction on levels and level morphisms.} In Agda 2.6.1, there is an
internal type of finite levels, and while construction rules and some built-in
operations on levels are exposed, there is no general elimination rule on
levels. Thus, there is a $\Nat \to \Lvl$ conversion function but it has no
inverse. In contrast, TTFL supports arbitrary elimination on levels and
morphisms.

\textbf{Type formers returning in least upper bounds of levels}. It is common in
type theories to allow type formers to have parameter types in different
universe levels, say $i$ and $j$, and return in level $i \sqcup j$. In TTFL,
whenever levels are \emph{trichotomous}, meaning that the ordering and equality
of levels is internally decidable, $i \sqcup j$ can be defined as the greater of
$i$ and $j$, and the ``heterogeneous'' type formers are derivable\footnote{A
level structure which is trichotonomous and supports \emph{extensionality},
i.e.\ $(\forall i.\, (i < j) \iff (i < k)) \to j = k$, is a
\emph{type-theoretic ordinal}. Assuming excluded middle, type-theoretic
ordinals are equivalent to classical ordinals \cite[Section 10.3]{hottbook}.}.

\textbf{Coercive cumulative subtyping.}
TTFL as specified does not directly support cumulative subtyping. However, it is
compatible with coercive subtyping. Consider the following rules:
\begin{alignat*}{3}
  &\blank\!\leq\!\blank &&: \Ty\,\Gamma\,i \to \Ty\,\Gamma\,j \to \Seti\\
  &\msf{coerce} &&: A \leq B \to \Tm\,\Gamma\,A \to \Tm\,\Gamma\,B\\
  &\leq\!\msf{refl} &&: A \leq A \\
  &\U\!\leq    &&: i < i' \to \U\,i\,j\,p \leq \U\,i'\,k\,q\\
  &\Pi\!\leq   &&: (p : A' \leq A) \to ((a' : \Tm\,\Gamma\,A')
  \to B[x \mapsto \msf{coerce}\,p\,a'] \leq B'[x \mapsto a'])\\
  & && \hspace{1em} \to \Pi(x : A) B \leq \Pi (x : A') B'
\end{alignat*}
Any model of TTFL can support the above rules: we can define
$\blank\!\leq\!\blank$ and $\msf{coerce}$ by indexed induction-recursion
\cite{indexedir}, where we define coercion along $\U\!\leq$ by type lifting, and
coercion along $\Pi\!\leq$ by backwards-forwards coercion.

Note that $\Pi$ is contravariant in the domain. This is easily supported
with our IR semantics, unlike in the set-theoretic model of cumulativity for Coq
\cite{timany2018cumulative}, where function domains are invariant.

\section{Related Work} \label{sec:related}

\section{Future Work and Conclusion} \label{sec:conclusion}
